Draft

### 2019-10-21 draft.tex
UTT, MP, CT, LLPO consistent

\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{axiom}[thm]{Axiom}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{definition}
\newtheorem{rem}[thm]{Remark}
\newtheorem{ill}[thm]{Illustration}
\newtheorem{exmp}[thm]{Example}
\crefname{thm}{Lemma}{Lemmas}

@@

	Given mere proposition $B$ and $f : A \rightarrow B$, the recursion principle gives us $g : \|A\| \rightarrow B$ such that $g(|a|) \equiv f(a)$ for all $a : A$.
	This recursion principle will prove itself an indispensable tool in the sections to come. 
	Whenever, in the midst of a proof, the current goal $B$ is a mere proposition and we have access to some witness $a : \lVert A \rVert$, we are allowed by the recursion principle to assume that we have $a' : A$ and use it to construct a proof of $B$.
	\item We have an \textbf{empty} type $\bot : \mathcal{U}$ with the induction principle
	$$\mathtt{ind}_\bot : \prod_{A : \mathcal{U}}\prod_{C : A \rightarrow \mathcal{U}} \prod_{x  :\bot} C\;x$$
	\item We have a unit type $\mathbf{1} : \mathcal{U}$ for which we have a term $* : \mathbf{1}$ and a witness of $\prod_{x,y : \mathbf{1}} x =y$. 
	\item We have the type of \textbf{natural numbers} $\mathbb{N} : \mathcal{U}$ equiped with $0 : \mathbb{N}$ and $\mathtt{succ} : \mathbb{N} \rightarrow \mathbb{N}$. 
	There is also the induction principle
	$$\mathtt{ind}_\mathbb{N} : \prod_{C : \mathbb{N} \rightarrow \mathcal{U}}C(0)\rightarrow \big(\prod_{ : \mathbb{N}} C(n) \rightarrow C(\mathtt{succ}\;n)\big)\rightarrow \prod_{n : \mathbb{N}}C\;n$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, 0) :\equiv f_0$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, \mathtt{succ}\;n) :\equiv f_s\big(n, \mathtt{ind}_\mathbb{N}(C,f_0, f_s, n)\big)$$

@@

We would also like a way to do classical logic in Type Theory. 
To do so we would need a way to say that some type $C$ is inhabited without providing a specific witness $c : C$, which would have the undesirable effect of providing more information than a classical statement of existence. 
An effective way to do so is to provide a witness of the truncation $\lVert C \rVert$ and say that $C$ is \textit{merely} inhabited. 
So to work classically we restrict ourselves to using only mere propositions. 
This approach is called `propositions as mere propositions' in \cite{hottbook}. 
\begin{itemize}
\item When we talk of conjunction $A \wedge B$, where $A$ and $B$ are mere propositions, we mean the product $A \times B$. 
\item We interpret $A \vee B$, where $A$ and $B$ are mere propositions, as the truncation $\| A + B\|$.
\item $\neg A$ is $A \rightarrow \bot$. 
\item We interpret $\forall a\in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\prod_{a : A}P(a)$. 
\item We interpret $ \exists a \in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\| \sum_{a : A} P(a) \|$
\end{itemize}
Note that the above are all chosen so that they preserve mere propositions, e.g. $A\times B$ is a mere proposition. 
The use of the above notation is interchangeable in the sections to follow. 
We will try to specify whether a type is \textit{merely} or \textit{actually} inhabited to avoid ambiguity whenever needed. 
\section{Formal Type Theory}
We shall now have a look at what Intensional Type Theory \textit{actually is}. 
We will give a short description of a formal system in which the informal derivations we do in the rest of this text are meant to be taking place. 
Building blocks of the formal system will be juxtaposed against their informal counterparts. 
We direct those looking for a full description to Appendixes A.1 and A.2 of \cite{hottbook}.

@@

The judgment $\Gamma \vdash a  : A$ corresponds to the exact step of an informal derivation where we construct the witness $a : A$ using terms over $\Gamma$ which we constructed in earlier steps of the same derivation. 
As one might expect, the deductive system is populated with inference rules that allow us to deduce valid judgments from collections of others; to illustrate
$$\frac{J_1,J_2,\ldots,J_n}{J}$$
All the constructions and rules described in \cref{informalTT} should and can be transcribed in this form.

@@

\begin{defn}
Given modality $\bigcirc : \mathcal{U} \rightarrow \mathcal{U}$, the \textbf{reflective subuniverse} of $\mathcal{U}$ is encoded by the following type
$$\mathcal{U}^\bigcirc \equiv \sum_{A : \mathcal{U}}\mathtt{isModal}_\bigcirc (A)$$
% \begin{itemize}
% 	\item For $A : \mathcal{U}$, we have $\mathtt{isModal}_{\bigcirc}(\bigcirc A)$.
% 	\item For $A : \mathcal{U}$ and $B$ such that $\mathtt{isModal}_{\bigcirc}(B)$, the function
% 	$$\lambda (f : \bigcirc A \rightarrow B).(f \circ \eta_A^\bigcirc) : (\bigcirc A \rightarrow B) \rightarrow (A \rightarrow B)$$
% 	is an equivalence.

% \end{itemize}
The subuniverse is $\Sigma$\textbf{-closed} if for $X$ such that $\mathtt{isModal}_{\bigcirc}(X)$ and $Q : X \rightarrow \mathcal{U}$ such that $\prod_{x : X} \mathtt{isModal}_{\bigcirc}(Q(x))$, we have $\mathtt{isModal}_{\bigcirc}(\Sigma_{x : X} Q(x))$.
\end{defn}
The \textit{reflective} modifier refers to the fact that for each $f : A \rightarrow B$ there is a canonical way to construct its \textit{reflection}, $f' : \bigcirc A \rightarrow \bigcirc B$. 
We simply compose $f$ with the modal unit $\eta_B^\bigcirc$ and use the induction principle of the modality on the result.

@@

\begin{thm}
Reflective subuniverses are closed under dependent products.
That is, for the subuniverse of $\bigcirc$ and $B : A \rightarrow \mathcal{U}$ such that\\
$\prod_{a : A}\mathtt{isModal}_{\bigcirc}(B(a))$, we have that $\mathtt{isModal}_{\bigcirc}(\prod_{a : A} B(a))$. 
\end{thm}

@@

\begin{lem}\label{405}
Under Church's thesis, the following type is inhabited
$$\sum_{F : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbf{2}} \Big( \big(\prod_{m : \mathbb{N}} \mathtt{atMost1one}\; F(m) \big) \times \big( \prod_{f : \mathbb{N} \rightarrow \mathbf{2}} \big\lVert \sum_{m, k : \mathbb{N}} F(m, 2k+ f\; m) = 1 \big\rVert \big) \Big)$$
\end{lem}
\begin{proof}
First we provide a witness
$$G : \sum_{F : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbf{2} } \prod_{n : \mathbb{N}}\mathtt{atMost1one}\; F(n)$$
For $n, m : \mathbb{N}$ we pick the indexed $T_n$ Turing machine of our enumeration. 
We can decide whether $m$ is odd or even. 
\begin{itemize}
\item If it's odd, then there exists actual $k : \mathbb{N}$ such that $m = 2k +1$. 
If $k$ is the Gödel number of the computation history that $T_n$ goes through when given $n$ as the input, furthermore, if it halts at the end of this computation with output $1$ then set $G.\mathtt{fst}\; n\; m :\equiv 1$. 
Otherwise set $G.\mathtt{fst}\; n\; m :\equiv 0$.
\item If it's even, then there exists actual $k$ such that $m = 2k$. 
We work as in the odd case, with the only difference being that we set $G.\mathtt{fst}\; n\; m :\equiv 1$ iff the output of the halting computation is $0$.
\end{itemize}
Having defined $G.\mathtt{fst}$ it's easy to see that $G.\mathtt{snd}$ has a straightforward proof, which we ommit.

We now define 
$$Q : (\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbf{2}) \rightarrow \mathcal{U}$$ 
by 
$$Q\; F :\equiv \prod_{f : \mathbb{N} \rightarrow \mathbf{2}} \big\lVert \sum_{m, k : \mathbb{N}} F(m, 2k+ f\; m) = 1 \big\rVert$$
We want to prove $Q\; G.\mathtt{fst}$. 
Consider arbitrary $f : \mathbb{N} \rightarrow \mathbf{2}$. 
By CT we have that there exists, merely, $e : \mathbb{N}$ such that $T_e$ computes $f$, furthermore there exists $z : \mathbb{N}$ which $z$ is the Gödel number of the halting computation that $T_e$ goes through when given $e$ as an input and lastly $T_e$ halts at the end of this computation and outputs $j : \mathbb{N}$ where $j = f\; e$. 
Since it's decidable whether $j$ is $0$ or $1$, we can take cases on it.
In both cases we have $G.\mathtt{fst} (e, 2z + j) = 1$.

@@

\begin{thm}
%In the context of HoTT, 
$$ACC \times CT \times LLPO \rightarrow \bot$$
\end{thm}
\begin{proof}
Let $G$ be the witness we reached in our proof of \cref{405}. 
Let $F :\equiv G.\mathtt{fst}$. 
By LLPO we can procure 
$$f : \prod_{n : \mathbb{N}} \lVert \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n) \rVert$$
By the ACC we get
$$f' : \big\lVert \prod_{n : \mathbb{N}}  \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n) \big\rVert$$
Since our goal is $\bot$, which is an h-prop, we can ignore the truncation and act as if we have access to
$$f'' : \prod_{n : \mathbb{N}}  \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n)$$
Let $e : \big( \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n)\big) \rightarrow \mathbf{2}$ be the function that sends $\mathtt{inl}\; \_ : \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n)$
to $1$ and $\mathtt{inr}\; \_$ to $0$. 
%By CT, the composition $e \circ f'' : \mathbb{N} \rightarrow \mathbf{2}$ has a $T_z$ that computes it. 
It's easy to see that the following holds
$$q : \prod_{n : \mathbb{N}} \big( (e\circ f''\; n = 1) \rightarrow \mathtt{p_{odd}}(F\; n) \big) \times \big( (e\circ f''\; n = 0) \rightarrow \mathtt{p_{even}}(F\; n) \big)$$
By $G.\mathtt{snd}$ we have that there exist $m,k : \mathbb{N}$ such that we have 
$$l : F(m, 2k + e\circ f''\; m) = 1$$
This should normally be truncated, but given the context, we are allowed to drop it.

@@

\chapter{Extensional Type Theory}
There exists a model of extensional type theory that validates the following `Independence Principle'
$$\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} P\; s \rightarrow \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbb{N}} Q\; s\; z\Big\rVert \bigg)$$
$$\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} P\; s \rightarrow  \Big\lVert \sum_{z:\mathbb{N}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow Q \; s\;z\Big\rVert \bigg)$$
where $P: \mathbf{2}^\mathbb{N} \rightarrow \mathtt{Prop}$ and $Q : \mathbf{2}^\mathbb{N} \rightarrow \mathbb{N}\rightarrow \mathtt{Prop} $ are families of propositions. 
Putting it plainly, if the left hand side of the above is true, then $z : \mathbb{N}$ does not depend on the proof of the constantness of $s$ to $0$.

The same model also validates Markov's Principle
$$\Big(\prod_{n : \mathbb{N}}(P\; n + \neg P\; n)\Big) \rightarrow  \Big(\neg \prod_{n : \mathbb{N}}\neg P\; n\Big)\rightarrow \Big\lVert \sum_{n : \mathbb{N}} P\; n \Big\rVert$$
where $P : \mathbb{N} \rightarrow \mathtt{Prop}$ is a family of propositions over $\mathbb{N}$. 
Informally, Markov's Principle states that if we have a collection of decidable propositions and the fact that not all of them are false, then one of them must be true.

In this section we work in the context of this model of Extensional Type Theory.
\begin{defn}
Given $f : C \rightarrow D$ we say that it's constant if for any $c_1,c_2 : C$, $f(c_1) = f(c_2)$.\\
Formally,
$$\mathtt{isConst}\; f :\equiv \prod_{c_1,c_2 : C}f(c_1) = f(c_2)$$
\end{defn}
Let $A :\equiv \sum_{a: \mathbf{2}^\mathbb{N}}\mathtt{atMost1one}\;a$. 
When referring to elements of $A$ we implicitly mean the first part of the pair. 
Let
$$B : A \rightarrow \mathcal{U}$$
$$B :\equiv \lambda\; a.\; \mathtt{p_{odd}}\; a + \mathtt{p_{even}}\; a$$

We would like to have access to the following instance of IP in our intensional model as well.
\begin{multline*}
\prod_{a: A} \prod_{h : \sum_{k : B\;a \rightarrow \mathbb{N}}\mathtt{isConst}(k)} \Bigg( 
\end{multline*}
\begin{multline*}
\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	 \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbb{N}}\prod_{p: B\; a}  k\; p = z \Big\rVert \bigg)
\end{multline*}
\begin{multline*}
\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	  \Big\lVert \sum_{z : \mathbb{N}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \prod_{p : B\; a} k\; p = z \Big\rVert \bigg) \Bigg)
\end{multline*}
The fact that $\mathbb{N}$ is $\lVert B \rVert$-null, follows from this in Intensional Type Theory. 
Our plan of action shall be to find a consequent of it, IP', strong enough for our purposes, that has the right form so that by Theorem 6.1 of \cite{1905.03014} we have a model of intensional type theory that satisfies IP'.

@@

$(\Leftarrow) $ Suppose $q : \prod_{a : A_1 + A_2} \lVert B\; a \rVert$. 
We want to construct a witness for $\prod_{a : A_1}\lVert B_1\; a\rVert \times \prod_{a : A_2}\lVert B_2\;a \rVert$. 
We do so by constructing witnesses for both of the constituents of the product. 
Wlog we do that only for $\prod_{a : A_1}\lVert B_1\; a\rVert$. 
Let $a : A_1$. Then $q(\mathtt{inl}\; a) : \lVert B_1\; a \rVert$, by the definition of $B$.
\end{proof}
\begin{cor}
The above lemma generalizes from the case of two families of types $B_1, B_2$ to a finite collection of families $B_1,\ldots,B_n$. 
\end{cor}
This fact enables us to reformulate a finite collection of statements, each one in the correct form, to a single statement in that same form which is equivalent to their conjunction.

We are now in the position to assume that we have access to a model $\mathcal{E}$ of univalent type theory, in which CT, MP and IP' hold.

>>>>

\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{axiom}[thm]{Axiom}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{definition}
\newtheorem{rem}[thm]{Remark}
\newtheorem{ill}[thm]{Illustration}
\newtheorem{exmp}[thm]{Example}

@@

	Given mere proposition $B$ and $f : A \rightarrow B$, the recursion principle gives us $g : \|A\| \rightarrow B$ such that $g(|a|) \equiv f(a)$ for all $a : A$.
	This recursion principle will prove itself an indispensable tool in the sections to come. 
	Whenever, in the midst of a proof, the current goal $B$ is a mere proposition and we have access to some witness $a : \lVert A \rVert$, we are allowed by the recursion principle to assume that we have $a' : A$ and use it to construct a proof of $B$.
	\item We have an \textbf{empty} type $\bot : \mathcal{U}$ with the induction principle
	$$\mathtt{ind}_\bot : \prod_{A : \mathcal{U}}\prod_{C : A \rightarrow \mathcal{U}} \prod_{x  :\bot} C\;x$$
	\item We have a unit type $\mathbf{1} : \mathcal{U}$ for which we have a term $* : \mathbf{1}$ and a witness of $\prod_{x,y : \mathbf{1}} x =y$. 
	\item We have the $\mathbf{2}$ type with terms $0 : \mathbf{2}$ and $1 : \mathbf{2}$ and the induction principle
	$$\mathtt{ind}_\mathbf{2} : \prod_{C : \mathbf{2} \rightarrow \mathcal{U}}C(0)\rightarrow C(1) \rightarrow \prod_{x : \mathbf{2}}C(x)$$
	$$\mathtt{ind}_\mathbf{2}()$$
	\item We have the type of \textbf{natural numbers} $\mathbb{N} : \mathcal{U}$ equiped with $0 : \mathbb{N}$ and $\mathtt{succ} : \mathbb{N} \rightarrow \mathbb{N}$. 
	There is also the induction principle
	$$\mathtt{ind}_\mathbb{N} : \prod_{C : \mathbb{N} \rightarrow \mathcal{U}}C(0)\rightarrow \big(\prod_{ : \mathbb{N}} C(n) \rightarrow C(\mathtt{succ}\;n)\big)\rightarrow \prod_{n : \mathbb{N}}C\;n$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, 0) :\equiv f_0$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, \mathtt{succ}\;n) :\equiv f_s\big(n, \mathtt{ind}_\mathbb{N}(C,f_0, f_s, n)\big)$$

@@

We would also like a way to do classical logic in Type Theory. 
To do so we would need a way to say that some type $C$ is inhabited without providing a specific witness $c : C$, which would have the undesirable effect of providing more information than a classical statement of existence. 
An effective way to do so is to provide a witness of the truncation $\lVert C \rVert$ and say that $C$ is \textit{merely} inhabited. 
So to work classically we restrict ourselves to using only mere propositions. 
This approach is called `propositions as mere propositions' in \cite{hottbook}. 
\begin{itemize}
\item When we talk of conjunction $A \wedge B$, where $A$ and $B$ are mere propositions, we mean the product $A \times B$. 
\item We interpret $A \vee B$, where $A$ and $B$ are mere propositions, as the truncation $\| A + B\|$.
\item $\neg A$ is $A \rightarrow \bot$. 
\item We interpret $\forall a\in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\prod_{a : A}P(a)$. 
\item We interpret $ \exists a \in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\| \sum_{a : A} P(a) \|$
\end{itemize}
Note that the above are all chosen so that they preserve mere propositions, e.g. $A\times B$ is a mere proposition. 
The use of the above notation is interchangeable in the sections to follow. 
We will try to specify whether a type is \textit{merely} or \textit{actually} inhabited to avoid ambiguity whenever needed. 
\section{Formal Type Theory}\label{formalTT}
We shall now have a look at what Intensional Type Theory \textit{actually is}. 
We will give a short description of a formal system in which the informal derivations we do in the rest of this text are meant to be taking place. 
Building blocks of the formal system will be juxtaposed against their informal counterparts. 
We direct those looking for a full description to Appendixes A.1 and A.2 of \cite{hottbook}.

@@

The judgment $\Gamma \vdash a  : A$ corresponds to the exact step of an informal derivation where we construct the witness $a : A$ using terms over $\Gamma$ which we constructed in earlier steps of the same derivation. 
As one might expect, the deductive system is populated with inference rules that allow us to deduce valid judgments from collections of others; to illustrate
$$\frac{J_1,J_2,\ldots,J_n}{J}$$
All the constructions and rules described in \ref{informalTT} should and can be transcribed in this form.

@@

\begin{defn}\label{reflectiveSubuniverse}
Given modality $\bigcirc : \mathcal{U} \rightarrow \mathcal{U}$, the $\sum$-closed \textbf{reflective subuniverse} of $\mathcal{U}$ is encoded by the following type
$$\mathcal{U}^\bigcirc \equiv \sum_{A : \mathcal{U}}\mathtt{isModal}_\bigcirc (A)$$
% \begin{itemize}
% 	\item For $A : \mathcal{U}$, we have $\mathtt{isModal}_{\bigcirc}(\bigcirc A)$.
% 	\item For $A : \mathcal{U}$ and $B$ such that $\mathtt{isModal}_{\bigcirc}(B)$, the function
% 	$$\lambda (f : \bigcirc A \rightarrow B).(f \circ \eta_A^\bigcirc) : (\bigcirc A \rightarrow B) \rightarrow (A \rightarrow B)$$
% 	is an equivalence.

% \end{itemize}
That the subuniverse is $\sum$\textbf{-closed} means that for $X$ such that $\mathtt{isModal}_{\bigcirc}(X)$ and $Q : X \rightarrow \mathcal{U}$ such that $\prod_{x : X} \mathtt{isModal}_{\bigcirc}(Q(x))$, we have $\mathtt{isModal}_{\bigcirc}(\Sigma_{x : X} Q(x))$.
\end{defn}
The \textit{reflective} modifier refers to the fact that for each $f : A \rightarrow B$ there is a canonical way to construct its \textit{reflection}, $f' : \bigcirc A \rightarrow \bigcirc B$. 
We simply compose $f$ with the modal unit $\eta_B^\bigcirc$ and use the induction principle of the modality on the result.

@@

\begin{thm}\label{reflSubuniversePiTypes}
Reflective subuniverses are closed under dependent products.
That is, for the subuniverse of $\bigcirc$ and $B : A \rightarrow \mathcal{U}$ such that\\
$\prod_{a : A}\mathtt{isModal}_{\bigcirc}(B(a))$, we have that $\mathtt{isModal}_{\bigcirc}(\prod_{a : A} B(a))$. 
\end{thm}

@@

\begin{lem}\label{405}
Under Church's thesis, the following type is inhabited
$$\sum_{F : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbf{2}} \Big( \big(\prod_{m : \mathbb{N}} \mathtt{atMost1one}\; F(m) \big) \times \big( \prod_{f : \mathbb{N} \rightarrow \mathbf{2}} \big\lVert \sum_{m, k : \mathbb{N}} F(m, 2k+ f\; m) = 1 \big\rVert \big) \Big)$$
\end{lem}
\begin{proof}
First we provide a witness
$$G : \sum_{F : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbf{2} } \prod_{n : \mathbb{N}}\mathtt{atMost1one}\; F(n)$$
For $n, m : \mathbb{N}$ we pick the indexed $T_n$ Turing machine of our enumeration. 
We can decide whether $m$ is odd or even. 
\begin{itemize}
\item If it's odd, then there exists actual $k : \mathbb{N}$ such that $m = 2k +1$. 
If $k$ is the G\"odel number of the computation history that $T_n$ goes through when given $n$ as the input, furthermore, if it halts at the end of this computation with output $1$ then set $G.\mathtt{fst}\; n\; m :\equiv 1$. 
Otherwise set $G.\mathtt{fst}\; n\; m :\equiv 0$.
\item If it's even, then there exists actual $k$ such that $m = 2k$. 
We work as in the odd case, with the only difference being that we set $G.\mathtt{fst}\; n\; m :\equiv 1$ iff the output of the halting computation is $0$.
\end{itemize}
Having defined $G.\mathtt{fst}$ it's easy to see that $G.\mathtt{snd}$ has a straightforward proof, which we ommit.

We now define 
$$Q : (\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbf{2}) \rightarrow \mathcal{U}$$ 
by 
$$Q\; F :\equiv \prod_{f : \mathbb{N} \rightarrow \mathbf{2}} \big\lVert \sum_{m, k : \mathbb{N}} F(m, 2k+ f\; m) = 1 \big\rVert$$
We want to prove $Q\; G.\mathtt{fst}$. 
Consider arbitrary $f : \mathbb{N} \rightarrow \mathbf{2}$. 
By CT we have that there exists, merely, $e : \mathbb{N}$ such that $T_e$ computes $f$, furthermore there exists $z : \mathbb{N}$ which $z$ is the G\"odel number of the halting computation that $T_e$ goes through when given $e$ as an input and lastly $T_e$ halts at the end of this computation and outputs $j : \mathbb{N}$ where $j = f\; e$. 
Since it's decidable whether $j$ is $0$ or $1$, we can take cases on it.
In both cases we have $G.\mathtt{fst} (e, 2z + j) = 1$.

@@

\begin{thm}
%In the context of HoTT, 
$$ACC \times CT \times LLPO \rightarrow \bot$$
\end{thm}
\begin{proof}
Let $G$ be the witness we reached in our proof of \ref{405}. 
Let $F :\equiv G.\mathtt{fst}$. 
By LLPO we can procure 
$$f : \prod_{n : \mathbb{N}} \lVert \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n) \rVert$$
By the ACC we get
$$f' : \big\lVert \prod_{n : \mathbb{N}}  \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n) \big\rVert$$
Since our goal is $\bot$, which is an h-prop, we can ignore the truncation and act as if we have access to
$$f'' : \prod_{n : \mathbb{N}}  \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n)$$
Let $e : \big( \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n)\big) \rightarrow \mathbf{2}$ be the function that sends $\mathtt{inl}\; \_ : \mathtt{p_{odd}}(F\; n) + \mathtt{p_{even}}(F\;n)$
to $1$ and $\mathtt{inr}\; \_$ to $0$. 
%By CT, the composition $e \circ f'' : \mathbb{N} \rightarrow \mathbf{2}$ has a $T_z$ that computes it. 
It's easy to see that the following holds
$$q : \prod_{n : \mathbb{N}} \big( (e\circ f''\; n = 1) \rightarrow \mathtt{p_{odd}}(F\; n) \big) \times \big( (e\circ f''\; n = 0) \rightarrow \mathtt{p_{even}}(F\; n) \big)$$
By $G.\mathtt{snd}$ we have that there exist $m,k : \mathbb{N}$ such that we have 
$$l : F(m, 2k + e\circ f''\; m) = 1$$
This should normally be truncated, but given the context, we are allowed to drop it.

@@

\chapter{UTT + CT + MP + LLPO}
In this chapter we present the main result of this text, which is showing that Univalent Type Theory is consistent with Church's thesis, Markov's principle and LLPO. 
Markov's Principle is the following statement
$$\Big(\prod_{n : \mathbb{N}}(P\; n + \neg P\; n)\Big) \rightarrow  \Big(\neg \prod_{n : \mathbb{N}}\neg P\; n\Big)\rightarrow \Big\lVert \sum_{n : \mathbb{N}} P\; n \Big\rVert$$
where $P : \mathbb{N} \rightarrow \mathtt{Prop}$ is a family of propositions over $\mathbb{N}$. 
Informally, Markov's Principle states that if we have a collection of decidable propositions and the fact that not all of them are false, then one of them must be true.

To reach the consistency result, we will work with models and results presented in \cite{1905.03014}. 
First, some quick terminology. 
A cwf (category with families) model of type theory is a categorical construction which `realizes' all data of formal TT presented in \ref{formalTT}. 
A detail of their construction that is of interest to us, is that any type over some context is `realized' in the cwf model as the set of its terms over the same context.

A quick overview of the overarching proof goes like this:\\
Start with a model of ETT+CT+MP. 
Show that this model satisfies a statement that will go by the name of IP'. 
Construct, based on this model, another model of UTT+CT+MP+IP'. 
Identify the null - with respect to LLPO - types of this model to retrieve a model of UTT+CT+MP+LLPO.\\
We proceed with the actual proof.

As the first step towards our goal, we consider the cwf model built on top of category of internal cubical objects in $\mathbf{Asm}(\mathcal{K}_1)$. 
For the category of assemblies over Kleene's first model, $\mathbf{Asm}(\mathcal{K}_1)$, we direct the reader to Chapter 1 of \cite{vanOosten:2008:RVI:1816956}. 
We call this model $\mathcal{E}$. 
By Theorem 3.10 of \cite{1905.03014} $\mathcal{E}$ satisfies Assumption 3.1 of the same paper. 
So it is a model of extensional type theory. 
Furthermore, Church's Thesis and Markov's Principle hold in it, as shown in the proof of theorem 6.4.  of \cite{1905.03014}. \todo{should i argue more about this?}

In addition to the above, $\mathcal{E}$ validates the following `Independence Principle' \todo{i claim, but don't I have to rationalize this?}
$$\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} P\; s \rightarrow \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbb{N}} Q\; s\; z\Big\rVert \bigg)$$
$$\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} P\; s \rightarrow  \Big\lVert \sum_{z:\mathbb{N}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow Q \; s\;z\Big\rVert \bigg)$$
where $P: \mathbf{2}^\mathbb{N} \rightarrow \mathtt{Prop}$ and $Q : \mathbf{2}^\mathbb{N} \rightarrow \mathbb{N}\rightarrow \mathtt{Prop} $ are families of propositions. 
Putting it plainly, if the left hand side of the above is true, then $z : \mathbb{N}$ does not depend on the proof of the constantness of $s$ to $0$.

%In this section we work in the context of this model of Extensional Type Theory.
\begin{defn}
Given $f : C \rightarrow D$ we say that it's constant if for any $c_1,c_2 : C$, $f(c_1) = f(c_2)$.\\
Formally,
$$\mathtt{isConst}\; f :\equiv \prod_{c_1,c_2 : C}f(c_1) = f(c_2)$$
\end{defn}
\begin{rem}\label{AandB}
Let $A :\equiv \sum_{a: \mathbf{2}^\mathbb{N}}\mathtt{atMost1one}\;a$. 
When referring to elements of $A$ we implicitly mean the first part of the pair. 
Let
$$B : A \rightarrow \mathcal{U}$$
$$B :\equiv \lambda\; a.\; \mathtt{p_{odd}}\; a + \mathtt{p_{even}}\; a$$
\end{rem}
We set as our new subgoal to reach an Orton-Pitts model, as seen in Chapter 3 of \cite{1905.03014}, which models UTT, CT and MP. 
We call this model to be constructed $\mathcal{E}'$. 
Furthermore, we require that $\mathbb{N}$ is $\lVert B \rVert$-null in this model. 
To make sure that this last requirement is met, we would like the following instance of IP to hold in it. 
\begin{multline*}
\prod_{a: A} \prod_{h : \sum_{k : B\;a \rightarrow \mathbb{N}}\mathtt{isConst}(k)} \Bigg( 
\end{multline*}
\begin{multline*}
\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	 \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbb{N}}\prod_{p: B\; a}  k\; p = z \Big\rVert \bigg)
\end{multline*}
\begin{multline*}
\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \\
	  \Big\lVert \sum_{z : \mathbb{N}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \prod_{p : B\; a} k\; p = z \Big\rVert \bigg) \Bigg)
\end{multline*}
The fact that $\mathbb{N}$ is $\lVert B \rVert$-null, follows from this in Intensional Type Theory. 
Our plan of action shall be to compromise and find a consequent of it, IP', weaker than IP in ETT but strong enough to imply $\lVert B \rVert$-nullness of $\mathbb{N}$ in UTT, that has the right form so that by Theorem 6.1 of \cite{1905.03014} we get our $\mathcal{E}'$ that satisfies IP'. 
Note that we have been and will be working in ETT up until the construction of $\mathcal{E}'$ is finalized. 
This is because we are proving results pertinent to $\mathcal{E}$.

@@

$(\Leftarrow) $ Suppose $q : \prod_{a : A_1 + A_2} \lVert B\; a \rVert$. 
We want to construct a witness for $\prod_{a : A_1}\lVert B_1\; a\rVert \times \prod_{a : A_2}\lVert B_2\;a \rVert$. 
We do so by constructing witnesses for both of the constituents of the product. 
Wlog we do that only for $\prod_{a : A_1}\lVert B_1\; a\rVert$. 
Let $a : A_1$. Then $q(\mathtt{inl}\; a) : \lVert B_1\; a \rVert$, by the definition of $B$.
\end{proof}
\begin{cor}
The above lemma generalizes from the case of two families of types $B_1, B_2$ to a finite collection of families $B_1,\ldots,B_n$. 
\end{cor}
This fact enables us to reformulate a finite collection of statements, each one in the correct form, to a single statement in that same form which is equivalent to their conjunction. 
So, given that CT, MP and IP' are in the right form required by Theorem 6.1 of \cite{1905.03014}, by the corollary above and said theorem, we have that there exists Orton-Pitts model $\mathcal{E}'$ of UTT, CT, MP and IP'.
\begin{prop}\label{EPrime}
$\mathcal{E}'$ is an Orton-Pitts model of UTT in which CT, MP and IP' hold.
\end{prop}
